BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-01-27 paper/paper.qmd
Redefine tensor operations and add to Lipschitz constant section

### Operations

The Frobenius inner product between two tensors $A,B\in\mathbb{R}^{I_1\times\dots\times I_N}$ is denoted

@@

@@

For matrices $M$, the (Operator) 2-norm is defined as

$$
\left\lVert M \right\rVert_2 = \arg\,\max_{\left\lVert v \right\rVert_2 = 1} \left\lVert Mv \right\rVert_2 = \sigma_1(M)
$$

@@

For tensors $T$, the (Operator) 2-norm needs to be defined in terms of how we treat them as function on other tensors. There is a canonical way to do this for vectors $x \to v^{\top}x$ and matrices $x \to Mx$, but not tensors. This is relevant to @sec-lipschitz-computation where the Lipschitz step-size is computed in terms of the Operator norm of the Hessian of our objective function.

@@

$$
Y = B\times_1 A_1 \times_2 \dots\times_N A_N = B \bigtimes_{n} A_n = \llbracket B; A_1, \dots, A_N \rrbracket.
$$

@@

@@

### Computing Gradients
- Use Auto diff generally
- But hand-crafted gradients and Lipschitz calculations *can* be faster (e.g. symmetrized slicewise dot product)

@@

@@

```julia
function make_gradient(T::Tucker1, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    if n==0 # the core is the zeroth factor
        function gradient0(T::Tucker1; kwargs...)
            (B, A) = factors(T)
            AA = A'A
            YA = Y×₁A'
            grad = B×₁AA - YA
            return grad
        end
        return gradient0
    elseif n==1 # the matrix is the first factor
        function gradient1(T::Tucker1; kwargs...)
            (B, A) = factors(T)
            BB = slicewise_dot(B, B)
            YB = slicewise_dot(Y, B)
            grad = A*BB - YB
            return grad
        end
        return gradient1
    else
        error("No $(n)th factor in Tucker1")
    end
end
```

@@

and the gradient with respect to the matrix factor $A_m$ is

$$
\nabla_{A_m} f(B, A_1, \dots, A_N) = A_m (B \bigtimes_{n\neq m} A_n) \cdot_m (B \bigtimes_{n\neq m} A_n) - Y \cdot_m (B \bigtimes_{n\neq m} A_n).
$$

```julia
function make_gradient(T::Tucker, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n==0 # the core is the zeroth factor
        function gradient_core(T::AbstractTucker; kwargs...)
            C = core(T)
            matrices = matrix_factors(T)
            gram_matrices = map(A -> A'A, matrices) # gram matrices AA = A'A, BB = B'B...
            YAB = tuckerproduct(Y, adjoint.(matrices)) # Y ×₁ A' ×₂ B' ...
            grad = tuckerproduct(C, gram_matrices) - YAB
            return grad
        end
        return gradient_core

    elseif n in 1:N # the matrix factors start at m=1
        function gradient_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n)
            An = factor(T, n)
            grad = An*slicewise_dot(TExcludeAn, TExcludeAn; dims=n) - slicewise_dot(Y, TExcludeAn; dims=n)
            return grad
        end
        return gradient_matrix

@@

```julia
function make_gradient(T::CPDecomposition, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n in 1:N # the matrix factors start at m=1
        function gradient_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n)
            An = factor(T, n)
            grad = An*slicewise_dot(TExcludeAn, TExcludeAn; dims=n) - slicewise_dot(Y, TExcludeAn; dims=n)
            return grad
        end
        return gradient_matrix

@@

The function `tuckerproduct(B, (A₁, ..., Aₙ))` computes
$$
B \bigtimes_{n} A_n = \llbracket B; A_1,\dots,A_N \rrbracket,
$$

and can optionally "exclude" one of the matrix factors `tuckerproduct(B, (A₁, ..., Aₙ); exclude=m)` to compute

$$
B \bigtimes_{n\neq m} A_n = \llbracket B; A_1,\dots,A_{n-1}, \text{id}_{I_n}, A_{n+1},\dots,A_{N} \rrbracket
$$
where $B\in\mathbb{R}^{I_1\times \dots \times I_N}$.

@@

# Computational Techniques

## For Improving Convergence Speed
- As stated, algorithm works
- But can be slow, especially for constrained or large problems

@@

@@

# Partial Projection and Rescaling
- for bounded linear constraints
  - first project
  - then rescale to enforce linear constraints
- faster to execute then a projection
- often does not loose progress because of the rescaling (decomposition dependent)

# Multi-scale
- use a coarse discretization along continuous dimensions
- factorize
- linearly interpolate decomposition to warm start larger decompositions

>>>>

### Products of Tensors

The outer product $\otimes$ between two tensors $A\in\mathbb{R}^{I_1\times\dots\times I_M}$ and $B\in\mathbb{R}^{J_1\times\dots\times J_N}$ yields an order $M+N$ tensor $A\otimes B \in \mathbb{R}^{I_1\times\dots\times I_M\times J_1\times\dots\times J_N}$ that is entry-wise

$$
(A\otimes B)[i_1,\dots,i_M,j_1,\dots,j_N] = A[i_1,\dots,i_M] B[j_1,\dots,j_N].
$$

TODO Define in BlockTensorDecomposition.jl

The Frobenius inner product between two tensors $A,B\in\mathbb{R}^{I_1\times\dots\times I_N}$ yields a real number $A \cdot B\in\mathbb{R}$ and is defined as

@@

The $n$-mode product and $n$-slice product can be thought of as opposites of each other. The $n$-mode product sums over just the $n$th dimension of the first tensor, whereas the $n$-slice product sums over all but the $n$th dimension.

We can extend the $n$-mode product to sum over multiple indices between two tensors.

The (complete) multi-mode product $\times_{1,\dots,n}=\times_{[n]}$ between a tensor $A\in\mathbb{R}^{I_1 \times \dots \times I_N}$ and tensor $B\in\mathbb{R}^{I_1\times\dots\times I_n}$, returns a tensor $(A \times_{[n]} B)\in\mathbb{R}^{I_{n+1} \times \dots \times I_N}$ with entries

$$
(A \times_{[n]} B)[i_{n+1},\dots,i_N] = \sum_{i_1, \dots ,i_n} A[i_1, \dots,i_n,i_{n+1},\dots,i_N]B[i_1, \dots ,i_n].
$$

### Gradients, Norms, and Lipschitz Constants

The gradient $\nabla f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}^{I_1\times \dots \times I_N}$ of a (differentiable) function $f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}$ is defined entry-wise for a tensor $A\in \mathbb{R}^{I_1\times \dots \times I_N}$ by

$$
\nabla f(A) [i_1, \dots, i_N] = \frac{\partial f}{\partial A[i_1, \dots, i_N]} (A).
$$

The Hessian $\nabla^2 f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}^{(I_1\times \dots \times I_N)^2}$ of a (second-differentiable) function $f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}$ is the gradient of the gradient and is defined for a tensor $A\in\mathbb{R}^{I_1\times \dots \times I_N}$ entry-wise by

$$
\nabla^2 f(A) [i_1, \dots, i_N, j_1, \dots , j_N] = \frac{\partial^2 f}{\partial A[i_1, \dots , i_N]\partial A[j_1, \dots , j_N]} (A).
$$

For a tensor input $A$ of order $N$, the Hessian tensor $\nabla^2 f(A)$ is of order $2N$.

@@

For matrices $M$, the (operator) 2-norm is defined as

$$
\left\lVert M \right\rVert_{\text{op}} = \left\lVert M \right\rVert_2 = \sup_{\lVert v \rVert_2 = 1} \left\lVert Mv \right\rVert_2 = \sigma_1(M)
$$

@@

For tensors $T$, the (operator) 2-norm needs to be defined in terms of how we treat them as function on other tensors. There is a canonical way to do this for vectors $x \to v^{\top}x$ and matrices $x \to Mx$, but not tensors. This is relevant to @sec-lipschitz-computation where the Lipschitz step-size is computed in terms of the Operator norm of the Hessian of our objective function. In the case of the Hessian tensor $\nabla^2 f(A)$ evaluated at $A$, we consider the function $X \to \nabla^2 f(A) \times_{[N]} X$. This gives us

$$
\left\lVert \nabla^2 f(A) \right\rVert_{\text{op}} = \sup_{\left\lVert X \right\rVert_F = 1} \left\lVert \nabla^2 f(A) \times_{[N]} X \right\rVert_F = \sigma_1(\text{mat}(\nabla^2 f(A)))
$$

where $\text{mat}:\mathbb{R}^{(I_1\times \dots \times I_N)^2}\to\mathbb{R}^{(I_1 \dots I_N)\times (I_1 \dots I_N)}$ takes a $2N$-ordered tensor and flattens it into a square $(I_1 \dots I_N)\times (I_1 \dots I_N)$ matrix.

A function $g:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}^{I_1\times \dots \times I_N}$ is $L$-Lipschitz when

$$
\left\lVert g(A) - g(B) \right\rVert_F \leq L \left\lVert A - B \right\rVert_F, \quad \forall A,B\in\mathbb{R}^{I_1\times \dots \times I_N}.
$$

We call the smallest such $L$ *the* Lipschitz constant of $g$.

A differentiable function $f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}$ is $L$-smooth when its gradient $g=\nabla f$ is $L$-Lipschitz.

::: {#thm-quadratic-smoothness}
# Quadratic Smoothness
Let $f:\mathbb{R}^{I_1\times \dots \times I_N}\to \mathbb{R}$ be a quadratic function

$$
f(X) = \frac{1}{2}A(X,X) + B(X) + C
$$

of $X$ with bilinear function $A:(\mathbb{R}^{I_1\times \dots \times I_N})^2\to\mathbb{R}$, linear function $B:\mathbb{R}^{I_1\times \dots \times I_N}\to\mathbb{R}$, and constant $C\in\mathbb{R}^{I_1\times \dots \times I_N}$.

Then

1. the Hessian $\nabla^2 f$ is a constant function that evaluates to $\nabla^2 f(X) = D$ at every point $X$ for some $D \in \mathbb{R}^{(I_1\times \dots \times I_N)^2}$,
2. the tensor $D$ only depends on the bilinear function $A$ (and not on $B$ and $C$), and
2. the quadratic function $f$ is $L$-smooth with constant

$$
L = \left\lVert D \right\rVert _{\text{op}} = \sigma_1(\operatorname{mat}D).
$$

@@

$$
Y = B\times_1 A_1 \times_2 \dots\times_N A_N = B \bigtimes_{n} A_n = \llbracket B; A_1, \dots, A_N \rrbracket.
$$ {#eq-tucker-product}

@@

The *Tucker Product* defined by @eq-tucker-product is implemented in BlockTensorDecomposition.jl with `tuckerproduct(B, (A1, ..., AN))` and computes
$$
B \bigtimes_{n} A_n = \llbracket B; A_1,\dots,A_N \rrbracket.
$$

It can also optionally "exclude" one of the matrix factors with the call `tuckerproduct(B, (A1, ..., AN); exclude=n)` to compute

$$
B \bigtimes_{m\neq n} A_m = \llbracket B; A_1,\dots,A_{n-1}, \text{id}_{R_n}, A_{n+1},\dots,A_{N} \rrbracket.
$$

```julia
function tuckerproduct(core, matrices; exclude=nothing)
    N = ndims(core)
    if isnothing(exclude)
        return multifoldl(tucker_contractions(N), (core, matrices...))
    else
        return multifoldl(getnotindex(tucker_contractions(N), exclude), (core, getnotindex(matrices, exclude)...))
    end
end

tucker_contractions(N) = Tuple((B, A) -> nmode_product(B, A, n) for n in 1:N)
```

@@

### Computing Gradients {#sec-gradient-computation}
- Use Auto diff generally
- But hand-crafted gradients and Lipschitz calculations *can* be faster (e.g. symmetrized slicewise dot product)

@@

::: callout-note
For the family of Tucker decompositions, the objective function $f$ is "block-quadratic" with respect to the factors. This means the gradient with respect to a factor is an affine function of that factor. This is exactly what we see in @eq-tucker-1-gradient-2 where $B$ is multiplied by the "slope" $A^\top A$ plus a shift of $-Y\times_1 A^\top$.
:::

@@

```julia
function make_gradient(T::Tucker1, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    if n==0 # the core is the zeroth factor
        function gradient0(X::Tucker1; kwargs...)
            (B, A) = factors(X)
            AA = A'A
            YA = Y×₁A'
            grad = B×₁AA - YA
            return grad
        end
        return gradient0
    elseif n==1 # the matrix is the first factor
        function gradient1(X::Tucker1; kwargs...)
            (B, A) = factors(X)
            BB = slicewise_dot(B, B)
            YB = slicewise_dot(Y, B)
            grad = A*BB - YB
            return grad
        end
        return gradient1
    else
        error("No $(n)th factor in Tucker1")
    end
end
```

@@

and the gradient with respect to the matrix factor $A_n$ is

$$
\nabla_{A_n} f(B, A_1, \dots, A_N) = A_n ( \tilde{X}_n \cdot_n \tilde{X}_n) - Y \cdot_n \tilde{X}_n
$$

where

$$
\tilde{X}_n = (B \bigtimes_{m\neq n} A_m) = \llbracket B; A_1,\dots,A_{n-1}, \text{id}_{R_n}, A_{n+1},\dots,A_{N} \rrbracket.
$$

```julia
function make_gradient(T::Tucker, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n==0 # the core is the zeroth factor
        function gradient_core(X::AbstractTucker; kwargs...)
            B = core(X)
            matrices = matrix_factors(X)
            gram_matrices = map(A -> A'A, matrices) # gram matrices AA = A'A,
                                                    # BB = B'B, ...
            grad = tuckerproduct(B, gram_matrices)
                 - tuckerproduct(Y, adjoint.(matrices))
            return grad
        end
        return gradient_core

    elseif n in 1:N # the matrix factors start at m=1
        function gradient_matrix(X::AbstractTucker; kwargs...)
            B = core(X)
            matrices = matrix_factors(X)
            Aₙ = factor(X, n)
            X̃ₙ = tuckerproduct(B, matrices; exclude=n)
            grad = Aₙ * slicewise_dot(X̃ₙ, X̃ₙ; dims=n)
                   - slicewise_dot(Y, X̃ₙ; dims=n)
            return grad
        end
        return gradient_matrix

@@

For the CP Decomposition, we can simply treat the core as $B=\text{id}_R$ and compute the gradient with respect to the matrix factors similarly to the Tucker decomposition:

$$
\nabla_{A_n} f(A_1, \dots, A_N) = A_n (\tilde{X}_n \cdot_n \tilde{X}_n) - Y \cdot_n \tilde{X}_n
$$

where

$$
\tilde{X}_n = (\text{id}_R \bigtimes_{m\neq n} A_m) = \llbracket \text{id}_R; A_1,\dots,A_{n-1}, \text{id}_{R}, A_{n+1},\dots,A_{N} \rrbracket.
$$

```julia
function make_gradient(T::CPDecomposition, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n in 1:N # the matrix factors start at m=1
        function gradient_matrix(X::AbstractTucker; kwargs...)
            B = core(X)
            matrices = matrix_factors(X)
            Aₙ = factor(X, n)
            X̃ₙ = tuckerproduct(B, matrices; exclude=n)
            grad = Aₙ * slicewise_dot(X̃ₙ, X̃ₙ; dims=n)
                   - slicewise_dot(Y, X̃ₙ; dims=n)
            return grad
        end
        return gradient_matrix

@@

@@

For the family of Tucker decompositions, we can compute the Lipschitz constants efficiently similar to how we compute the gradient in @sec-gradient-computation.

::: callout-note
For the family of Tucker decompositions, the objective function $f$ is "block-quadratic" with respect to the factors. We saw in @eq-tucker-1-gradient-2 that the gradient is an affine map where $B$ is multiplied by the "slope" $A^\top A$ plus a shift of $-Y\times_1 A^\top$.

This means the Lipschitz constant of the gradient for that block is the operator norm of the "slope".
:::

Since our least-squares objective $f$ is a block-quadratic function in each factor, we can use @thm-quadratic-smoothness to find the smoothness $L_{A_n}$ with respect to each factor $A_n$ by calculating the largest eigenvalue of the partial Hessian $\nabla^2_{A_n}f(A_1,\dots,A_N)$.

Starting with the Tucker-1 decomposition (@def-tucker-1-decomposition) objective $f(B,A)=\frac{1}{2}\left\lVert AB-Y\right\rVert_F^2$, we have

$$
\nabla^2_{B}f(B,A) = A
$$

TODO Explain how we get the following. General idea is to take partial Hessian of our quadratic function with respect to one block, and get that the operator norm of the hessian (as a map from $\mathbb{R}^{I_1\times \dots \times I_N}$ to itself) is the Lipschitz constant of the gradient. See Obsidian note "Gradients Hessians and Lipschitz Constants of a Tucker Decomposition".

```julia
function make_lipschitz(T::Tucker1, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    if n==0 # the core is the zeroth factor
        function lipschitz0(T::Tucker1; kwargs...)
            A = matrix_factor(T, 1)
            return opnorm(A'A)
        end
        return lipschitz0

    elseif n==1 # the matrix is the zeroth factor
        function lipschitz1(T::Tucker1; kwargs...)
            C = core(T)
            return opnorm(slicewise_dot(C, C))
        end
        return lipschitz1

    else
        error("No $(n)th factor in Tucker1")
    end
end

function make_lipschitz(T::Tucker, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n==0 # the core is the zeroth factor
        function lipschitz_core(T::AbstractTucker; kwargs...)
            #matrices = matrix_factors(T)
            #gram_matrices = map(A -> A'A, matrices)
            #return prod(opnorm.(gram_matrices))
            return prod(A -> opnorm(A'A), matrix_factors(T))
        end
        return lipschitz_core

    elseif n in 1:N # the matrix is the zeroth factor
        function lipschitz_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n)
            return opnorm(slicewise_dot(TExcludeAn, TExcludeAn; dims=n))
        end
        return lipschitz_matrix

    else
        error("No $(n)th factor in Tucker")
    end
end

function make_lipschitz(T::CPDecomposition, n::Integer, Y::AbstractArray; objective::L2, kwargs...)
    N = ndims(T)
    if n in 1:N # the matrix is the zeroth factor
        function lipschitz_matrix(T::AbstractTucker; kwargs...)
            matrices = matrix_factors(T)
            TExcludeAn = tuckerproduct(core(T), matrices; exclude=n) # TODO optimize this to avoid making the super diagonal core
            return opnorm(slicewise_dot(TExcludeAn, TExcludeAn; dims=n))
        end
        return lipschitz_matrix

# Computational Techniques
- As stated, algorithm works
- But can be slow, especially for constrained or large problems

As stated, the algorithm described in @sec-base-algorithm works. It will converge to a solution to our optimization problem and factorize the input tensor. It is worth discussing how the algorithm can be modified to improve convergence to maintain quick convergence for large problems, and what sort of architectural methods are used to allow for maximum flexibility, without over engineering the package.

## For Improving Convergence Speed

There are a few techniques used to assist convergence. Two ideas that are well studied are discussed in this section. They are 1) breaking up the updates into smaller blocks, and 2) using momentum or acceleration. What is perhaps novel is considering the synergy between these two ideas.

Two more techniques are implemented in BlockTensorDecomposition.jl to improve convergence. To the authors knowledge, these are new to tensor factorization, but may or may not be applicable depending on the exact factorization problem or data being studied. For these reasons, these other techniques are discussed separately in @sec-ppr and @sec-multi-scale.

@@

When using block coordinate descent as in @sec-base-algorithm, it is natural to treat each factor as its own block. This requires the fewest blocks while ensuring the objective is still convex with respect to each block. We could just as easily use smaller blocks.

In [@lee_algorithms_2000], the multiplicative update algorithm used to solve nonnegative matrix factorization problems is written in a way so that each row of the first factor

@@

# Partial Projection and Rescaling {#sec-ppr}
- for bounded linear constraints
  - first project
  - then rescale to enforce linear constraints
- faster to execute then a projection
- often does not loose progress because of the rescaling (decomposition dependent)

# Multi-scale {#sec-multi-scale}
- use a coarse discretization along continuous dimensions
- factorize
- linearly interpolate decomposition to warm start larger decompositions
